---
title: "Genomic Data Visualization using R"
author: "Jubayer Hossain"
date: "23 February, 2021"
output:
  html_document:
    toc: true
    toc_depth: 6
    toc_float: true
    theme: flatly
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment=NA)
```

## What is ggplot2?
- ggplot2 is a data visualization package for the statistical programming language R.
- Created by Hadley Wickham in 2005, ggplot2 is an implementation of Leland Wilkinson’s Grammar of Graphics—a general scheme for data visualization which breaks up graphs into semantic components such as scales and layers.

## Types of Visualization
In statistics, we generally have two kinds of visualization:

- Exploratory data visualization: Exploring the data visually to find patterns among the data entities.
- Explanatory data visualization: Showcasing the identified patterns using simple graphs.

## Why Visualization?
“A picture is worth a thousand words”

- Data visualizations make big and small data easier for the human brain to understand, and visualization also makes it easier to detect patterns, trends, and outliers in groups of data.
- Good data visualizations should place meaning into complicated datasets so that their message is clear and concise.

## Grammar of Graphics
- Data - The dataset being plotted
- Aesthetics- The scales onto which we plot our data
- Geometry - The visual elements used for our data
- Facet	-Groups by which we divide the data

## Load Packages and Set Theme
```{r, pkg, message=FALSE}
library(tidyverse) 
library(gridExtra) 
theme_set(theme_gray(base_size = 18))
```

## Load Data 
```{r}
data <- read.csv("variants_from_assembly.bed", sep="\t", stringsAsFactors = TRUE, header = FALSE, quote = '')
```

## Exploring Data 
```{r, head}
# examine first few rows 
head(data) 
```

```{r, colnames}
# Colnames 
names(data) <-  c("chrom", "start", "stop", "name", "size", "strand", "type", "ref.dist", "query.dist")
```

```{r, check}
head(data) 
```

```{r, ds}
# Data Structures
glimpse(data)
```

```{r}
# Summary 
summary(data)
```

## Bar Chart 
```{r}
ggplot(data, aes(x=chrom))+
  geom_bar()
```

```{r}
# Filtering Data 
data <- filter(data, chrom %in% c(seq(1,22), "X", "Y", "MT"))
ggplot(data, aes(x=chrom))+
  geom_bar()
```

```{r}
# Ordering Chromosomes
data$chrom <- factor(data$chrom, levels=c(seq(1,22),"X","Y","MT"))
ggplot(data, aes(x=chrom))+
  geom_bar()
```

```{r}
# Xlab, Ylab, Title
ggplot(data, aes(x=chrom))+
  geom_bar()+
  labs(title = "Chromosomes Distribution", 
       x="Chromosomes", 
       y= "Counts")
```


```{r}
# Centering title 
ggplot(data, aes(x=chrom))+
  geom_bar()+
  labs(title = "Chromosomes Distribution",
       x="Chromosomes", 
       y= "Counts")+
  theme(plot.title = element_text(hjust = 0.5))
```


```{r}
# Color by categorical variable 
ggplot(data, aes(x=chrom, fill=type))+
  geom_bar()+
  labs(title = "Chromosomes Distributions", 
       x="Chromosomes", 
       y= "Counts")+
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
# Ordering type 
data$type <- factor(data$type, levels = c("Insertion","Deletion",                          "Expansion","Contraction"))
```


```{r}
# Color by categorical variable 
ggplot(data, aes(x=chrom, fill=type))+
  geom_bar()+
  labs(title = "Chromosomes Distributions", 
       x="Chromosomes", 
       y= "Counts")+
  theme(plot.title = element_text(hjust = 0.5))
```


```{r}
# Color by categorical variable 
ggplot(data, aes(x=chrom, fill=type))+
  geom_bar()+
  labs(title = "Chromosomes Distributions", 
       x="Chromosomes", 
       y= "Counts")+
  theme(plot.title = element_text(hjust = 0.5))+
  guides(fill=guide_legend(title = "Type"))
```

```{r}
fig <- function(width, heigth){
     options(repr.plot.width = width, repr.plot.height = heigth)
}
```


```{r}
# Facet Wrap 
ggplot(data, aes(x=chrom, fill=type))+
  geom_bar()+
  labs(title = "Chromosomes Distributions", 
       x="Chromosomes", 
       y= "Counts")+
  guides(fill=guide_legend(title = "Type"))+
  facet_wrap(~type)+
  fig(20,6)
```


```{r}
# Facet Grid 
ggplot(data, aes(x=chrom, fill=type))+
  geom_bar()+
  labs(title = "Chromosomes Distributions", 
       x="Chromosomes", 
       y= "Counts")+
  guides(fill=guide_legend(title = "Type"))+
  facet_grid(type ~ .)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

## Scatter Plot 

```{r}
ggplot(data, aes(x=ref.dist, y=query.dist))+
  geom_point()
```

```{r}
ggplot(data, aes(x=ref.dist, y=query.dist, color=type))+
  geom_point()+
  xlim(-500, 500)+
  ylim(-500, 500)
```

## Boxplot 

```{r}
ggplot(data, aes(x=type, y=size))+
  geom_boxplot()

```
```{r}
ggplot(data, aes(x=type, y=size))+
  geom_boxplot()+
  coord_flip()
```

## Violin Plot 

```{r}
ggplot(data, aes(x=type, y=size))+
  geom_violin()
```
```{r}
ggplot(data, aes(x=type, y=size))+
  geom_violin()+
  coord_flip()
```
```{r}
ggplot(data, aes(x=type, y=size))+
  geom_violin(adjust=0.2)+
  scale_y_log10()+
  coord_flip()
```


## Desnsity Plot 
```{r}
ggplot(data, aes(x=size, fill=type))+
  geom_density(alpha=0.5)+
  xlim(0,500)
```
```{r}
ggplot(data, aes(x=size, fill=type))+
  geom_density(alpha=0.5)+
  xlim(0,500)+
  facet_wrap(~type)
```

```{r}
ggplot(data, aes(x=size, fill=type))+
  geom_density(alpha=0.5)+
  xlim(0,500)+
  facet_grid(type ~ .)
```
```{r}
ggplot(data, aes(x=size, fill=type))+
  geom_density(alpha=0.5)+
  xlim(0,500)+
  facet_grid(. ~type)
```
```{r}

ggplot(data, aes(x=size, fill=type))+
  geom_density(alpha=0.5)+
  xlim(0,500)+
  facet_grid(type ~chrom)
```

## Pie Chart 
```{r}
type_counts <- summary(data$type)
pie(type_counts)
```


## Arrange Plot 
```{r}
f1 <- ggplot(data, aes(x=chrom,fill=type))+ 
  geom_bar()+
  labs(title = "Distribution of Chromosomes",
       x="Chromosomes", y="Counts", 
       tag = "A")+
  guides(fill=guide_legend(title = "Type"))

f2 <- ggplot(data, aes(x=size, fill=type))+
  geom_density(alpha=0.5)+
  xlim(0,500)+
  labs(title = "Density Plot of Size", 
       x = "Size", 
       y = "Probability", 
       tag = "B")+
  guides(fill=guide_legend(title = "Type"))


  grid.arrange(f1, f2, nrow=2, ncol=1)
```
